Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-07-31 improved_state_space.tex
finished reviewing ms. Now to the response letter

\subsection*{Robustness Simulation Study}
In simulations of the geometric-recruitment model estimated with exponential model,
estimates of $\Lambda$ were biased high %(2.71, relative bias 27.1\%) 
and $p$ biased low (Table A5). % (-0.05, relative bias -21.1\%).  
%As a result, RMSE for these parameters were high %(2.88 and 0.05, respectively) 
%and coverages low. % (13.0\% and 3.9\%).
However, estimates of $r$ had low bias %(-0.0005) 
and RMSE, %(0.005)
with nominal coverage. % (95.7\%).  
The estimates from the geometric-recruitment model for $\Lambda$ and $p$ had low bias, but the estimates
for $\gamma$ and $\omega$ were somewhat biased and imprecise (Table A5 and Fig. A3).
The geometric- recruitment model had lower
AIC 71.3\% of the time; when the geometric-recruitment model was favored, the
mean $\Delta$AIC was 6.0 and when the exponential model was favored
the mean was 1.4.  %Results from the Leslie matrix model simulations were similar, but more dramatic.
For the Leslie matrix simulations, bias in estimates of $\Lambda$ and $p$ from both the exponential and geometric-recruitment models
were high, but the exponential biases were higher (Table A6 and Fig. A3).  
The geometric-recruitment model was favored in 100\% of the simulations, 
with a mean $\Delta$AIC of 89.0.

@@

Estimates for the top ranked model 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved (Appendix B, Fig. B1). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
and the estimate for r was intermediate. The 
estimate of the intercept for $p$ dropped. 
When environmental stochasticity was added, the estimate for $K$ increased. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. 
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided little evidence of lack of convergence for any of the Bayesian models ran.

@@

One strategy for adhering to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data. For example, in a
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with data
from, for example,
capture-recapture data. Models that are fitted to count data and
demographic data are often referred to as integrated population models
\citep[IPM;][]{besbeas_etal:2002,
buckland_etal:2004,schaub_etal:2007}.
We believe that a DM modeling framework could work well in this context,
and that the DM model could be viewed as a specific case of the
IPM proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.

@@

Fortunately, incorporating direct information about detection probability 
is straightforward and we recommend that this be done whenever
possible. \citet{dail_madsen:2011} proposed that
replicated counts be conducted within an interval in which the populations can 
be assumed to be closed. A robust design \citep{pollock:1982} could be 
used to combine multiple surveys per
primary period to increase the precision of the estimates. We envision
that multiple other options are available as well, such as removal,
multiple observer, and distance modeling \citep{williams_etal:2002}. These could be
accomplished by extending the open population $N$-mixture model
in exactly the same way as the closed population version has been
extended \citep[e.g.,][]{royle_etal:2004}.

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually interesting
extensions that we believe would be computationally prohibitive in many
cases. Nonetheless, we will discuss one---spatially-explicit models of
immigration. 
In our models, immigration is currently modeled as independent of
population sizes in other sites. Several alternatives suggest
themselves if we think about movement in a metapopulation
context \citep{hanski:1998}: immigration conditional on total or mean
abundance across sites the previous time step; immigration
conditional on abundance in nearby sites only; and immigration
conditional on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
be based on the assumptions that either the study sites
encompass the whole range of the metapopulation or that the
annual changes in the populations sampled are representative of
a larger metapopulation. Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework.

>>>>

\subsection*{Robustness Simulation Study}
In simulations of the geometric-recruitment model estimated with the exponential model,
estimates of $\Lambda$ were biased high %(2.71, relative bias 27.1\%) 
and $p$ biased low (Table A5). % (-0.05, relative bias -21.1\%).  
%As a result, RMSE for these parameters were high %(2.88 and 0.05, respectively) 
%and coverages low. % (13.0\% and 3.9\%).
However, estimates of $r$ had low bias %(-0.0005) 
and RMSE, %(0.005)
with nominal coverage. % (95.7\%).  
The estimates from the geometric-recruitment model for $\Lambda$ and $p$ had low bias, but the estimates
for $\gamma$ and $\omega$ were somewhat biased and imprecise (Table A5 and Fig. A3).
The geometric- recruitment model had lower
AIC 71.3\% of the time; when the geometric-recruitment model was favored, the
mean $\Delta$AIC was 6.0 and when the exponential model was favored
the mean was 1.4.  %Results from the Leslie matrix model simulations were similar, but more dramatic.
For the Leslie matrix simulations, bias in estimates of $\Lambda$ and $p$ from both the exponential and geometric-recruitment models
were high, but the exponential biases were higher (Table A6 and Fig. A3).  
The geometric-recruitment model was favored in 100\% of the simulations, 
with a mean $\Delta$AIC of 89.0.

@@

Estimates for the top ranked model 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved (Appendix B, Fig. B1). %% What's up with that??
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically
and the estimate for r was intermediate. The 
estimate of the intercept for $p$ dropped. 
When environmental stochasticity was added, the estimate for $K$ increased. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. 
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided little evidence of lack of convergence for any of the Bayesian models ran.

@@

One strategy for adhering to the original formulation of the
model that included survival and recruitment parameters would be to
use either prior information or additional data \citep{zipkin_etal:2014}. For example, in a 
Bayesian analysis existing estimates of demographic parameters
could be used as informative priors, which might make it possible to
seperate the effects of movement versus births and deaths.
Alternatively, ecologists could combine count data with data
from, for example,
capture-recapture data. Models that are fitted to count data and
demographic data are often referred to as integrated population models
\citep[IPM;][]{besbeas_etal:2002,schaub_etal:2007,chandler_clark:2014}.
We believe that a DM modeling framework could work well in this context,
and that the DM model could be viewed as a specific case of the
IPM proposed by \citet{buckland_etal:2004}.
However, we caution that when count data are affected by
movement and vital rates, it would be unwise to ignore the movement
processes.

@@

Fortunately, incorporating direct information about detection probability 
is straightforward and we recommend that this be done whenever
possible. \citet{dail_madsen:2011} proposed that
replicated counts be conducted within an interval in which the populations can 
be assumed to be closed. A robust design \citep{pollock:1982} could be 
used to combine multiple surveys per
primary period to increase the precision of the estimates. We envision
that multiple other options are available as well, such as removal,
multiple observer, and distance sampling \citep{williams_etal:2002}. These could be
accomplished by extending the open population $N$-mixture model
in exactly the same way as the closed population version has been
extended \citep[e.g.,][]{royle_etal:2004}.

Our emphasis was on increasing the practical utility of this
class of models, and so we avoided several conceptually interesting
extensions that we believe would be computationally prohibitive in many
cases. Nonetheless, we will discuss one---spatially-explicit models of
immigration. 
In our models, immigration is currently modeled as independent of
population sizes in other sites. Several alternatives suggest
themselves if we think about movement in a metapopulation
context \citep{hanski:1998}: immigration conditional on total or mean
abundance across sites the previous time step; immigration
conditional on abundance in nearby sites only; and immigration
conditional on abundance at sites as a function of their distance
from the receiving site \citep{hastings:1991,hanski:1998}. These models would
require that all sites in the metapopulation are sampled, or that abundance at 
unsampled locations can be estimated \citep{lele_etal:1998}. 
%be based on the assumptions that either the study sites
%encompass the whole range of the metapopulation or that the
%annual changes in the populations sampled are representative of
%a larger metapopulation. 
Initial tests suggest that it would be
possible to fit some or all of these models, at least in the
Bayesian framework.
